Weakly Nonlinear Analysis of the Standard and Helical Magnetorotational Instabilities in a Taylor Couette Flow

### 2016-07-29 paper/widegap_mri_paper.tex
more intro on experiments

Our investigation is astrophysically motivated, but we also intend our theory to be relevant to laboratory experiments. Several experimental efforts are attempting to observe the MRI in the laboratory, which would allow the study of a crucial astrophysical phenomenon in a controlled setting. Unfortunately, detection of the MRI has so far proven elusive. \citei{Sisan:2004ig} claimed to detect the MRI, but most likely detected unrelated MHD instabilities instead (\citei{Hollerbach:2009ig}, \citei{Gissinger:2011td}). Most relevant to our work is the Princeton Plasma Physics Laboratory (PPPL) MRI Experiment, a liquid gallium Taylor-Couette flow exposed to an axial magnetic field (\citei{Ji:2001kd}).

@@

Because we include curvature terms, our theory is also relevant to the helical magnetorotational instability (HMRI). Discovered by \citeauthor{Hollerbach:2005tr} (\citeyear{Hollerbach:2005tr}), the HMRI is an overstability in which the background magnetic field is helical. The HMRI currently occupies a special place in the MRI puzzle. The HMRI has been proposed as a method of awakening angular momentum transport in the ``dead zones" of protoplanetary disks where the magnetic Prandtl number ($Pm = \nu/\eta$) becomes very small. However the rotation profiles needed to excite HMRI may be prohibitively steeper than Keplerian depending on the boundary conditions, and so its role in astrophysical disks is currently debated (\citei{Liu:2006}, \citei{Rudiger:2007}, \citei{Kirillov:2013}). The HMRI is significantly easier to excite in a laboratory setting than the standard MRI, and has already been detected in the laboratory by the Potsdam Rossendorf Magnetic Instability Experiment (PROMISE; \citei{Stefani:2006iv}, \citei{Stefani:2009hp}). The standard MRI has yet to be detected in the laboratory, although several plasma experiments are making progress.

@@

\bibitem[{Gissinger et al.(2011)}]{Gissinger:2011td}
Gissinger, C, Ji, H, Goodman, J, Phys. Rev. E, 2011, 84, 026308

@@

>>>>

Our investigation is astrophysically motivated, but we also intend our theory to be relevant to laboratory experiments. Several experimental efforts are attempting to observe the MRI in the laboratory, which would allow the study of a crucial astrophysical phenomenon in a controlled setting. Unfortunately, detection of the MRI has so far proven elusive. \citei{Sisan:2004ig} claimed to detect the MRI in a spherical Couette flow, but most likely detected unrelated MHD instabilities instead (\citei{Hollerbach:2009ig}, \citei{Gissinger:2011td}). Most relevant to our work is the Princeton Plasma Physics Laboratory (PPPL) MRI Experiment, a liquid gallium Taylor-Couette flow exposed to an axial magnetic field (\citei{Ji:2001kd}). Some theoretical work designed to complement the Princeton MRI experiment involves direct simulation of the experimental conditions, much of it focused on the specific challenges in identifying MRI signatures despite apparatus-driven spurious flows (e.g. \citei{Gissinger:2012gc}). In particular, the vertical endcaps on a laboratory MRI apparatus drives meridional flows which both inhibit the excitement of MRI and obscure its detection. The Princeton MRI experiment employs split, independently rotating endcaps to mitigate these flows (\citei{Schartman:2009df}). Our work here assumes an infinite vertical domain, an idealization that is theoretically expedient but experimentally impractical. However, our setup is designed to be an accurate treatment of the radial dimension of the flow in a Taylor Couette apparatus like the one used in the Princeton MRI experiment.

@@

Because we include curvature terms, our theory is also relevant to the helical magnetorotational instability (HMRI). Discovered by \citeauthor{Hollerbach:2005tr} (\citeyear{Hollerbach:2005tr}), the HMRI is an overstability in which the background magnetic field is helical. The HMRI currently occupies a special place in the MRI puzzle. The HMRI has been proposed as a method of awakening angular momentum transport in the ``dead zones" of protoplanetary disks where the magnetic Prandtl number ($Pm = \nu/\eta$) becomes very small. However the rotation profiles needed to excite HMRI may be prohibitively steeper than Keplerian depending on the boundary conditions, and so its role in astrophysical disks is currently debated (\citei{Liu:2006}, \citei{Rudiger:2007}, \citei{Kirillov:2013}). The HMRI is significantly easier to excite in a laboratory setting than the standard MRI, and has already been detected in the laboratory by the Potsdam Rossendorf Magnetic Instability Experiment (PROMISE; \citei{Stefani:2006iv}, \citei{Stefani:2009hp}). %The standard MRI has yet to be detected in the laboratory, although several plasma experiments are making progress.

@@

\bibitem[{Gissinger et al.(2011)}]{Gissinger:2011td}
Gissinger, C, Ji, H, Goodman, J, 2011, Phys. Rev. E 84, 026308

\bibitem[{Gissinger et al.(2012)}]{Gissinger:2012gc}
Gissinger, C, Goodman, J, \& Ji, H, 2012, Physics of Fluids 24, 074109

@@

\bibitem[{Schartman et al.(2009)}]{Schartman:2009df}
Schartman, E, Ji, H, Burin, M J, Rev Sci Instrum, 2009, 80, 024501
